---
title: "Introduction to the GenomicRanges package \n MonBUG presentation"
author: "Mathieu.Lajoie@gmail.com"
date: "Febuary 3, 2016"
output:
  ioslides_presentation:
    css: monbug.css
  slidy_presentation: default
---

```{r,include=FALSE}

  #SET YOUR WORKING DIRECTORY HERE
  setwd("/Users/lajoimat/Projects/GIT/lajoie_sinnett_lab/MonBug/")

  #INSTALL PACKAGES IF NEEDED
  #source("https://bioconductor.org/biocLite.R")
  #biocLite("GenomicRanges")
  #biocLite("TxDb.Dmelanogaster.UCSC.dm3.ensGene")
  #biocLite("pasillaBamSubset")

```

## Introduction

+ GenomicRanges is a Bioconductor package defining general purpose containers and functions for manipulating genomic intervals and associated annotations

+ Time and memory efficient (can handle high-throughput sequencing data)

+ Many R packages dedicated to genomic analysis rely on this package

+ If you do your downstream analysis with R, it can spare you from going back and forward between RStudio, Perl/Python/Bedtools scripts and your intermediate results files

+ Authors: P. Aboyoun, H. Pagès, and M. Lawrence


## GRanges object

A GRanges object is a *vector* of genomic intervals and associated annotations. 

Each element (interval) in the vector is comprised of

 + a sequence name (e.g. chr1)
 + a range (start & end)
 + a strand (+, -, *)
 + optional metadata columns (e.g. score, GC content, etc.).

## Creating GRanges objects{.codefont}
Lets start by creating a GRanges with a single interval.

```{r,message=FALSE}
library(GenomicRanges)
``` 
```{r}
gr <- GRanges(seqnames = "chr1",
              ranges = IRanges(50,100), 
              strand = "+")
gr
```

## Creating GRanges objects{.codefont}
When providing multiple ranges, *seqnames* and *strand* are recycled by the constructor if they are not vectors.
```{r}
gr <- GRanges(seqnames = "chr1",
              ranges = IRanges( 10:101, 15:106 ),strand = "-")
gr
```

## Rle{.codefont}

Rle is a simple form of lossless data compression in which runs of identical values are stored as (value ; run_length) pairs

```{r}
x = c(rep(1,50),rep(0,40),rep(1,30))
x
Rle(x)
```

## GRanges accessors{.codefont}

start(), end(), strand(), seqnames(), length()

```{r}
length(gr)
strand(gr)
start(gr)[1:20]
```


## GRanges annotations{.codefont}
Single *annotation* column can be set/get with $
```{r}
gr$score = 10:101
gr$dataset = "TP53_ChIPseq"
gr
```

## GRanges annotations{.codefont}
The annotations are stored in a *DataFrame* that can be accessed with *mcols()*
```{r}
mcols(gr)
```

## Combining GRanges objects{.codefont}
Just like with *base::vector*, we can use the *c()* function
```{r,message=FALSE}
gr.b = gr
gr.b$dataset = "MYC_ChIPseq"

gr.combined = c( gr, gr.b )
gr.combined
```

## Transformations on GRanges objects{.codefont}
We first create 3 GRanges objects each consisting of 20 intervals distributed around some locations (75, 200 and 275).
```{r,message=FALSE}

start.a = rnorm(n = 20, mean = 75, sd = 35) 
start.b = rnorm(n = 20, mean = 200, sd = 20)
start.c = rnorm(n = 20, mean = 275, sd = 15)

gr.a = GRanges("seq1", IRanges(start.a, start.a + 50), "+")
gr.b = GRanges("seq1", IRanges(start.b, start.b + 50), "-") # on minus strand
gr.c = GRanges("seq1", IRanges(start.c, start.c + 50), "+")

gr.abc = c(gr.a, gr.b, gr.c) #combined
```

## Simple plotting function{.codefont}
We define a simple plotting function
```{r}
gr.plot = function(gr,xlim=range(c(start(gr),end(gr))),...){
  #Find y lim for plotting region
  ylim = c(0,length(gr)+2)

  #Init empty plot
  plot(0,ylim=ylim,yaxt="n",ylab="", xlab="Genomic position",col="white", xlim=xlim, ...)
  if(length(gr)>0){
  #Draw ranges using rect()
  rect(start(gr), #left coordinate
       1:length(gr), #y bottom
       end(gr), #right coordinate
       2:(length(gr)+1),#y top
       col=1+as.numeric(as.factor(strand(gr))),
       border = "black")}
  legend("topleft",legend = c("+","-","*"),fill=2:4)
}
```

## Our 3 GRanges objects{.codefont}
```{r,message=FALSE}
gr.plot(gr.abc, xlim=c(50,450))
text(x=420,y=c(10,30,50),c("gr.a","gr.b","gr.c"),cex = 1.2)
abline(lty=2,h=c(21,41))
```

## Intra range transformations
These are applied to each range individually and the returned GRanges object has the same length as the input object

+ shift()
+ resize()

## Shifting GRanges coordinates{.codefont}
```{r}
gr.b2 = shift(gr.b,300)
```
```{r,echo=FALSE}
par(mfrow=c(1,2))
gr.plot(gr.abc, main = "original")
gr.plot(c(gr.a, gr.b2, gr.c), main = "shift(gr.b,300)")
```

## Resizing GRanges elements{.codefont}
```{r}
gr.b2 = resize(gr.b,200,"center")
```
```{r,echo=FALSE}
par(mfrow=c(1,2))
gr.plot(gr.abc,main="original")
gr.plot(c(gr.a, gr.b2, gr.c), main = "resize(gr.b, 200,\"center\")")
```

## Inter range transformations
These transforms all the ranges together as a set to produce a new set of ranges, returned in a GRanges object not necessarily of the same lenght.

+ reduce()
+ intersection()
+ union()

## GRanges reduce() {.codefont}
```{r}
gr.b2 = reduce(gr.b)
```
```{r,echo=FALSE}
par(mfrow=c(1,2))
gr.plot(gr.abc, main="original")
gr.plot(c(gr.a, gr.b2, gr.c),main="reduce(gr.b)")
```

## GRanges reduce() {.codefont}
```{r}
gr.ab = reduce(c(gr.a,gr.b))
```
```{r,echo=FALSE}
par(mfrow=c(1,2))
gr.plot(gr.abc, main="original")
gr.plot(c(gr.ab,gr.c),main="reduce(c(gr.a, gr.b))")
```

## GRanges reduce() {.codefont}
```{r}
gr.ab = reduce(c(gr.a,gr.b),ignore.strand=TRUE)
```
```{r,echo=FALSE}
par(mfrow=c(1,2))
gr.plot(gr.abc, main="original")
gr.plot(c(gr.ab,gr.c),main="reduce(c(gr.a, gr.b),\n ignore.strand = TRUE)")
```

## Intersect of 2 GRanges{.codefont}
```{r,message=FALSE}
gr.inter.ab = intersect(gr.a,gr.b)
```
```{r,echo=FALSE}
par(mfrow=c(1,2))
gr.plot(gr.abc, main = "original")
gr.plot(c(gr.inter.ab, gr.c), xlim = c(50,380), main="intersect(gr.a, gr.b)")
```

## Intersect of 2 GRanges{.codefont}
```{r,message=FALSE}
gr.inter.ab = intersect(gr.a,gr.b,ignore.strand = TRUE)
```
```{r,echo=FALSE}
par(mfrow=c(1,2))
gr.plot(gr.abc, main = "original")
gr.plot(c(gr.inter.ab, gr.c), xlim = c(50,380),main="intersect(gr.a, gr.b,\n ignore.strand==TRUE")
par(mfrow=c(1,1))
```

## Union of 2 GRanges {.codefont}
```{r}
gr.ab = union(gr.a,gr.b)
```
```{r,echo=FALSE}
par(mfrow=c(1,2))
gr.plot(gr.abc, main="original")
gr.plot(c(gr.ab,gr.c),main="union(gr.a, gr.b)")
```

## Finding overlapping GRanges{.codefont}
```{r}
overlapsAny(gr.b, gr.c) 
overlapsAny(gr.b, gr.c,ignore.strand=TRUE) 
```

## Finding overlapping GRanges{.codefont}
```{r}
gr.window = GRanges("seq1",IRanges(175,300),"*")
gr.overlapped = gr.abc[ overlapsAny(gr.abc, gr.window) ] 
```
```{r,echo=FALSE}
par(mfrow=c(1,2))
gr.plot(c(gr.window, gr.abc)); abline(v=c(175,300))
gr.plot(c(gr.window, gr.overlapped), xlim =c(50,max(end(gr.abc))), main="gr.overlapped") 
abline(v=c(175,300))
```

## coverage() function
```{r}
cov = coverage(gr.abc)
cov
```

## coverage() function
```{r,eval = FALSE}
plot(cov$seq1, type='l', lwd=2, main = "coverage gr.abc")
```
```{r,echo=FALSE}
par(mfrow=c(1,2))
gr.plot(gr.abc)
plot(cov$seq1, type='l', lwd=2, main = "coverage gr.abc")
```


## A simple function to read GRanges from MACS output file
```{r}
load_peak_file = function(file,header=FALSE){
  peaks.df <-read.delim(file,header=header, comment.char="#",
                        stringsAsFactors = FALSE)
  if(sum(grepl(pattern = "chr",peaks.df[,1]))==0){ 
    peaks.df[,1]=paste("chr",peaks.df[,1],sep="")
  }
  peaks=GRanges(seqnames=peaks.df[,1],ranges=IRanges(peaks.df[,2], peaks.df[,3])) 
  if (ncol(peaks.df) > 3) { 
      mcols(peaks) <- peaks.df[, 4:ncol(peaks.df)] 
  }
  peaks$filename = basename(file)
  peaks 
}
```

## Loading some real data (H3K4me3 peaks from ENCODE){.codefont}
```{r}
gr.H3K4me3 = load_peak_file("DATA/E116-H3K4me3.narrowPeak")
head(gr.H3K4me3, n=3)
```

## GRangesList object{.codefont}
```{r}
grl = split(gr.H3K4me3, seqnames(gr.H3K4me3))
names(grl)
grl$chr1
```

## Plotting peak distribution on chr1
```{r}
hist(start(grl$chr1),breaks=200,col="grey",
     xlab="chr1",main="peak distribution")
```

## Example with RNAseq data

Adapted from "An Introduction to Genomic Ranges Classes" available at https://bioconductor.org/

 + Read a bam file 
 + Load gene annotation
 + Get the read counts by gene
 + Plot read coverage for a specific gene
 + Get the number of reads supporting splice junctions
 
## Load appropriate packages

```{r, warning = FALSE, message=FALSE}

#Class for handling genomic alignments
library(GenomicAlignments) 

#Gene annotations for Dmelanogaster
library(TxDb.Dmelanogaster.UCSC.dm3.ensGene) 

#An RNAseq dataset
library(pasillaBamSubset) 

```

## Loading a BAM file as a GAlignements object
```{r}

# Path to the example BAM
un1 <- untreated1_chr4() 

gr.gal <- readGAlignments(un1) #Reading the BAM file

head(gr.gal,n=3) #Look at the first 3 alignments
```

## Extracting gene models 
```{r}

#Get a GRangesList containig the exons for each gene 
exbygene <- exonsBy(TxDb.Dmelanogaster.UCSC.dm3.ensGene, "gene")

exbygene
```


## Get the read counts by gene 
```{r}

# This function performs similarly to HTSeq-count
se <- summarizeOverlaps(exbygene, un1, mode="IntersectionNotEmpty")

# Extract count matrix from se object
counts = assays(se)$counts 

head(counts) #===> Ready for edgeR or DESeq analysis
```


## Get the gene model of a gene of interest
```{r}

# Lets take the one with the maximum number of counts
counts[which.max(counts[,1]),,drop=FALSE] # Well this is R...

# Get its gene model (list of exons)
gr.exons = exbygene[["FBgn0017545"]]

gr.exons
```

## Define a gene region for plotting purpose
```{r}

#Define the gene region
gr.gene = range(gr.exons) # min & max coordinate as a GRanges

#Define the exonic region
gr.exonic = reduce(gr.exons) #All the positions covered by at lest one exon
```

## Get the coverage for that region
```{r}
gr.cov = as(coverage(gr.gal),"GRanges") #Coerce Rle to GRanges

gr.cov = gr.cov[overlapsAny(gr.cov, gr.gene)] #Keep gene region
```

## Plot coverage & exonic region
```{r}
plot(start(gr.cov),gr.cov$score,type='l',lwd=5,xlab="chr4",ylab="coverage",col="grey")
rect(start(gr.exonic),-200,end(gr.exonic),200,col="lightgreen",lwd=3)
```

##Get the number of reads supporting splice junctions
```{r}
gr.junc = summarizeJunctions(gr.gal)
gr.sel.junc = gr.junc[overlapsAny(gr.junc, gr.gene)]
gr.junc # A GRanges
```

## Last plot
```{r,echo=FALSE}
plot(start(gr.cov),gr.cov$score,type='l',lwd=5,xlab="chr4",ylab="coverage",col="grey")
rect(start(gr.exonic),-200,end(gr.exonic),200,col="lightgreen",lwd=3)
points(start(gr.junc),gr.junc$score,pch=19,col="red")
points(end(gr.junc),gr.junc$score,pch=19,col="red")
legend("topleft",fill=c("grey","lightgreen","red"),
       legend = c("coverage","exonic regions","junction scores"))
```

## Time for Knitr ?

*knitr in a knutshell*, a very nice tutorial by Karl Broman

http://kbroman.org/knitr_knutshell


