---
title: "MonBUG presentation (practical stuff)"
author: "Mathieu.Lajoie@gmail.com"
date: "Febuary 3, 2016"
output:
  ioslides_presentation:
    css: monbug.css
  slidy_presentation: default
---

```{r,include=FALSE}
  library(DT)
  setwd("/Users/lajoimat/Projects/GIT/lajoie_sinnett_lab/MonBug/")
```

## Using serialization to save your data and intermediate results

Serialization is the process of translating data structures or object states into a stream of bytes so that it can be stored or transmitted and eventually reconstructed

+ Faster loading time
+ Smaller file size
+ No more time wasted on parsing headers and figuring out what was the field separator

## Example: ENCODE peak file for GM12878 H3K4me3
```{r}
peak_file = "DATA/E116-H3K4me3.narrowPeak"
df.peaks = read.table(peak_file)
```
```{r,echo=FALSE}
# File size (MB)
size= signif(file.size("DATA/E116-H3K4me3.narrowPeak")/1.05e6,digits=3)
print(paste("File size:",size,"MB"))
```

## Saving it using saveRDS()
Without compression
```{r}
saveRDS(df.peaks,"DATA/E116-H3K4me3.narrowPeak.rds")
```
```{r,echo=FALSE}
# File size (MB)
size= signif(file.size("DATA/E116-H3K4me3.narrowPeak.rds")/1.05e6,digits=3)
print(paste("File size:",size,"MB"))
```

With bzip2 compression (smaller but takes more time)
```{r}
saveRDS(df.peaks,"DATA/E116-H3K4me3.narrowPeak.rds",compress = "bzip2")
```
```{r,echo=FALSE}
# File size (MB)
size= signif(file.size("DATA/E116-H3K4me3.narrowPeak.rds")/1.05e6,digits=3)
print(paste("File size:",size,"MB"))
```

## Benchmarking
```{r}
library(rbenchmark)
benchmark(replications=rep(10),
           loading_from_txt = read.table(peak_file),
           loading_from_rds = readRDS("DATA/E116-H3K4me3.narrowPeak.rds"),
           columns=c('test', 'elapsed', 'replications'))
```

## Copy/Paste data in R {.codefont}
If you need to put data from an R table or vector into a spreadsheet or web form (for example gene identifiers), you can copy/paste directly in R (MacOSX/Linux version)

```{r}

write.clipboard <- function(x,row.names=FALSE,col.names=TRUE) {
  write.table(x,pipe("pbcopy"),sep="\t",row.names = row.names,col.names = col.names)
}

write.clipboard(df.peaks) # Example usage
```
```{r}
read.clipboard <- function(header = FALSE,row.names = NULL){
  read.table(pipe("pbpaste"), sep="\t", header = header, row.names = row.names)
}

df.new = read.clipboard() # Example usage
```

## DT package

```{r,message=FALSE,warning=FALSE}
library(DT)
datatable(df.peaks[1:500,1:6])
```

